---
title: "NOAA Storm Data Analysis"
author: "Tim Shores"
date: "July 25, 2015"
output: html_document
---
# Synopsis
Your data analysis must address the following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

- 10 sentences, no more.
- at least one figure containing a plot.
- no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.

# Data Processing
which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.

```{r load_data, echo = TRUE, cache = TRUE}
datapath <- 
    paste("/home/tim/datasciencecoursera/Practicing-R/Reproducible Research/PeerAssessment2/repdata_data_StormData.csv.bz2")
stormdata <- read.csv(datapath)
```

# Results
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

```{r, echo = TRUE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
