---
title: "NOAA Storm Data Analysis"
author: "Tim Shores"
date: "July 25, 2015"
output: html_document
---
# Synopsis
In this report we aim to identify storm events with the greatest impact on human health, and the greatest impact on valuable property, across the United States from January 1995 to November 2011.

The code chunks in this report will directly import the [U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2), and process the data to compute results relevant to the above questions. [Storm database documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) and [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf) are available.

# Data Processing

First we load software libraries and display the processing software environment.

```{r software_env, echo = TRUE}
library(dplyr)      # for cleaning and exploring data
library(ggplot2)    # for plots
options(scipen = 10) # to prevent numbers displayed with scientific notation
sessionInfo()
```

Next we load the data.

```{r load_data, echo = TRUE, cache = TRUE}
datapath <- 
    paste("/home/tim/datasciencecoursera/Practicing-R/Reproducible Research/PeerAssessment2/repdata_data_StormData.csv.bz2")
storm0 <- read.csv(datapath)
storm0$BGN_DATE <- as.POSIXct(storm0$BGN_DATE, format = "%m/%d/%Y %H:%M:%S")
    # This converts factor to dates, for easier filtering of rows.
```

After importing the data, we display details about the data frame, including the first few rows and columns. There are `r format(dim(storm0)[1], big.mark = ",")` rows and `r dim(storm0)[2]` columns in this dataset. The variable headers are descriptive enough, so I don't modify them.

```{r display_data, echo = TRUE}
dim(storm0)
head(storm0[, 1:8])
```

Next, we prepare data for the first question, which seeks to rank storm events by harmfulness to health. The question is not limited by timeframe or location. We only need variables for types of events and health outcomes, which we interpret to mean INJURIES and FATALITIES. These will be added together as a new variable, Harm.total. Since the NOAA database includes storm data beginning in 1950, we must filter out the earlier data.

```{r preprocess_question1, echo = TRUE}
storm1 <- 
    storm0 %>% 
    group_by(EVTYPE) %>% 
    select(EVTYPE, BGN_DATE, FATALITIES, INJURIES, REMARKS, REFNUM) %>%
    mutate(Harm.total = FATALITIES + INJURIES) %>%
    filter(BGN_DATE > "1994-12-31")
```

We also check for missing values in the selected dataset, coded as NA (there are none).

```{r checkNA1, echo = TRUE}
mean(is.na(storm1))
```

We next summarise this Harm.total data, grouped by storm event type, to obtain sum, mean, and standard deviation. We extract the top five sum and mean harmful events, then combine these uniquely to produce a set of most harmful storm events. We remove events with no standard deviation, as these indicate events that only happen once -- singular events should not attract the attention of managers.

```{r summarise_question1, echo = TRUE}
storm1.summary <- 
    storm1 %>% 
    summarise(Harm.sum = sum(Harm.total), 
              Harm.mean = round(mean(Harm.total), 5), 
              Harm.sd = sd(Harm.total)) %>% 
    filter(Harm.sum > 0)

# this creates three separate top five harmful sum, mean, and median, 
# then jams them together, removing duplicates.
top.harm.sum <- top_n(storm1.summary, 5, Harm.sum)
top.harm.mean <- top_n(storm1.summary, 5, Harm.mean)
storm1.top.harm <- unique(rbind(top.harm.sum, top.harm.mean))
storm1.top.harm <- filter(storm1.top.harm, !is.na(Harm.sd))

```

Next we prepare data for the second question, which seeks to rank storm events by economic consquences. This question is also not limited by timeframe or location. We only need variables for types of events and economic outcomes, which is coded as 'damage' values representing dollar amounts for property and for crops. 

Both damage variables are accompanied by DMGEXP factor variables that define the value magnitude: 

- K for thousands of dollars
- M for millions
- B for billions

I use the mutate() function in the first pipeline to normalize all damage values to thousands of dollars.

```{r preprocess_question2, echo = TRUE}
storm2 <- 
    storm0 %>% 
    group_by(EVTYPE) %>% 
    select(EVTYPE, BGN_DATE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, REMARKS, REFNUM) %>%
    mutate( 
       PROPDMG = ifelse(PROPDMGEXP == "K", 
                        PROPDMG, 
                        ifelse(PROPDMG == "M", 
                               PROPDMG * 1000, 
                               PROPDMG * 1000000)),
       CROPDMG = ifelse(CROPDMGEXP == "K",
                        CROPDMG,
                        ifelse(CROPDMG == "M",
                               CROPDMG * 1000,
                               CROPDMG * 1000000))
       ) %>%
    mutate(Dmg.total = PROPDMG + CROPDMG) %>%
    filter(BGN_DATE > "1994-12-31")
```

We also check for missing values in the selected dataset, coded as NA (again there are none).

```{r meanNA2, echo = TRUE}
mean(is.na(storm2))
```

As with question 1 harm data, the economic damage data is summarised by storm event type groups, producing a list of the top damaging event types that happen more than once.

```{r summarise_question2, echo = TRUE}
storm2.summary <- 
    storm2 %>%
    summarise(Dmg.sum = round(sum(Dmg.total), 2), 
              Dmg.mean = round(mean(Dmg.total), 2), 
              Dmg.sd = round(sd(Dmg.total), 2)) %>%
    filter(Dmg.sum > 0)

# this creates three separate top five damaging sum, mean, and median, 
# then jams them together, removing duplicates.
top.dmg.sum <- top_n(storm2.summary, 5, Dmg.sum)
top.dmg.mean <- top_n(storm2.summary, 5, Dmg.mean)
storm2.top.dmg <- unique(rbind(top.dmg.sum, top.dmg.mean))
storm2.top.dmg <- filter(storm2.top.dmg, !is.na(Dmg.sd))

```

# Results
Across the United States, this data frame shows events that are most harmful with respect to population health.

```{r harm_event_factors, echo = TRUE}
storm1.top.harm
```

This barplot compares the log of average harm per event. Error bars on each bar indicate standard deviation. Color fill indicates total damage of all incidents of an event type. 

```{r plot1_harm, echo = TRUE}
harm.p <- ggplot(data = storm1.top.harm, 
                 aes(x = EVTYPE, 
                     y = log(Harm.mean), 
                     fill = Harm.sum,
                     title = "Storm Health Harms 1995-2011"))
harm.p + geom_bar(stat = "identity") +
    xlab("Storm Event Type") + 
    theme(axis.text.x = element_text(angle=50, size=8)) +
    ylab("log Avg Injury/Fatality") + 
    scale_fill_continuous(name = "Injury/Fatality Count All Events") +
    geom_errorbar(aes(ymin = log(Harm.mean)-log(Harm.sd), ymax = log(Harm.mean)+log(Harm.sd)), 
                  width=.2, position=position_dodge(.9))
  
```

Across the United States, this data frame shows events that have the greatest economic consequences.

```{r damage_event_factors, echo = TRUE}
storm2.top.dmg
```

This barplot compares the log of average economic damage per event. Error bars on each bar indicate standard deviation. Color fill indicates total damage of all incidents of an event type. 

```{r plot2_damage, echo = TRUE}
dmg.p <- ggplot(data = storm2.top.dmg, 
                aes(x = EVTYPE, 
                    y = log(Dmg.mean), 
                    fill = Dmg.sum,
                    title = "Storm Economic Damage 1995-2011"))
dmg.p + geom_bar(stat = "identity") +
    xlab("Storm Event Type") + 
    theme(axis.text.x = element_text(angle=50, size=8)) +
    ylab("log Avg Damage (thousands USD)") + 
    scale_fill_continuous(name = "Total Damage All Events (thousands USD)") +
    geom_errorbar(aes(ymin = log(Dmg.mean)-log(Dmg.sd), ymax = log(Dmg.mean)+log(Dmg.sd)), 
                  width=.2, position=position_dodge(.9))
```